---
title: "Stat 420 Final Project Work"
author: "STAT 420, Summer 2022, D. Unger"
date: ''
output:
  pdf_document: default
  html_document: 
    theme: readable
    toc: yes  
urlcolor: cyan
---

# Analysis Workbook


```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
options(scipen = 1, digits = 4, width = 80, fig.alin = "center")

library(ggplot2)
```

## Proposal
7/10

Everyone make own file for work. 

## Define Helper Functions

## EDA
-shared by 7/13

```{r Kathryn EDA}
#Read in the data
housing_prices <- read.csv("C:/School/Stat_420/house_price_prediction/000_Data/california-housing-prices/housing.csv")

#review the first few rows
head(housing_prices)
str(housing_prices)
summary(housing_prices)

plot(housing_prices)

#Check ocean proximity
housing_prices$ocean_proximity <- as.factor(housing_prices$ocean_proximity)
table(housing_prices$ocean_proximity)

plot(log(median_house_value) ~ ocean_proximity, housing_prices)

#Check density per block 
boxplot(housing_prices$population / housing_prices$total_rooms)

#I think this obs should be thown out.
housing_prices[which.max(housing_prices$population / housing_prices$total_rooms),]
```


## Data cleaning
-agreed/coordinate
-shared by 7/13

```{r}
#TODO: Use lat/long to get zip code?
```


## Variable Creation
-shared 7/22

## Model Building
-shared 7/22

## Model Selection
- share 7/22
-coordinate 7/23

## Graphs and Tables  
- coordinate 7/13
-coordinate 7/23 

## QA: How do we know what we did makes sense?.

## Move to final report
August 2nd (remember to knit often!!)
Report Due August 5th
