---
title: "Stat 420 Final Project Proposal"
author: "STAT 420, Summer 2022, D. Unger"
date: ''
output:
  pdf_document: default
  html_document: 
    theme: readable
    toc: yes  
urlcolor: cyan
---

# Stat 420 Final Project Proposal

```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
options(scipen = 1, digits = 4, width = 80, fig.alin = "center")

library(ggplot2)
```

## 1. Team Member information

Team Member's Names: 

 - Kathryn DeWitt
 - Shashank Thakur
 - Avinash Tiwari

NetIDs: 

 - kdewitt3
 - sthakur5
 - tiwari6

## 2. Proposed Title: 

We propose calling our project **Modeling California Housing Prices**.

## 3. Description of Data File

'California Housing Prices' dataset pertains to the houses found in a given California district and some summary stats about them based on the 1990 census data. There are 10 columns and 20.6K rows in the data. The columns are as mentioned below: 

| **Column Name**    	| **Description**                                                                                   	|
|--------------------	|---------------------------------------------------------------------------------------------------	|
| longitude          	| Longitudinal location of block group of houses.                                                     |
| latitude           	| Latitudinal location of block group of houses                                                     	|
| housing_median_age 	| Median age of the houses in the block group                                                       	|
| total_rooms        	| Total rooms in the group of houses                                                                	|
| total_bedrooms     	| Total bedrooms in the block group                                                                 	|
| population         	| Total number of people in the block group                                                         	|
| households         	| Households comprised in the block group                                                           	|
| median_income      	| Median income calculated from the individual income of house.                                     	|
| median_house_value 	| Median house value of a block group.                                                              	|
| ocean_proximity    	| Indicating whether each block group is near the ocean, near the Bay Area, inland or on an island. 	|


## 4. Background on datasets

For our project we will be using the data set from kaggle [California Housing Prices](https://www.kaggle.com/datasets/camnugent/california-housing-prices)

The data contains information from the 1990 California census and pertains to houses found in a given California district. It contains one row per census block group. A block group is the smallest geographical unit for which the U.S. Census Bureau publishes sample data (a block group typically has a population of 600 to 3,000 people).


## 5. Research Question/Interest in Data Set

We would like to model median house value of a block group in CA in the 1990's based on administrative (Census) data. Given the recent turbulence in the real estate market and associated bidding wars, there's a renewed interest in what the value of a house is. The data set we propose using is interesting for the following reasons:

- Unlike scraped real estate data, the data set is gather by an administrative body, the Census. Since the stakeholder involved in collecting the data is not interested in selling a house, we see different features represented that relate more to geography (Latitude, longitude), population density (population, number of households), and age of the building (median_housing_age).
- The dataset is aggregated at the block level (rather than at the house level). This means our analysis is not directly modelling housing values, but the median housing value of a neighborhood. We  believe this will help us better to understand the value of the neighborhood as opposed to the value of the amenities of the individual houses.
- Additionally, this data set meets the requirements set out by the project description:
   - The dataset has more than 20,000 observations
   - The dataset has a good mix of categorical and numerical variables to enable us to learn from the course
   - The dataset has 10 columns
   - The dataset coming from Kaggle means there has been extensive work done by many researchers / learners around the world providing our team with a baseline to compare our model & methodology with others. We also learn from their work and explore how much improvement could be achieved by applying modeling techniques other than regression to achieve better model performance


## 6. Evidence of Loading into R

We have included below evidence that the data can be loaded into R.

```{r read_data_file}
ca_housing_data = read.csv('../000_Data/california-housing-prices/housing.csv')
head(ca_housing_data)
```



