---
title: "Stat 420 Final Project Proposal"
author: "STAT 420, Summer 2022, D. Unger"
date: ''
output:
  html_document: 
    theme: readable
    toc: yes  
  pdf_document: default
urlcolor: cyan
---

# Stat 420 Final Project Proposal

```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
options(scipen = 1, digits = 4, width = 80, fig.alin = "center")

library(ggplot2)
```

## 1. Team Member information

Team Member's Names: 

 - Kathryn DeWitt
 - Shashank Thakur
 - Avinash Tiwari

NetIDs: 

 - kdewitt3
 - sthakur5
 - tiwari6

## 2. Proposed Title: 

We propose calling our project **Modeling California Housing Prices**.

## 3. Description of Data File

'California Housing Prices' dataset pertains to the houses found in a given California district and some summary stats about them based on the 1990 census data. There are 10 columns and 20.6K rows in the data. The columns are as mentioned below: 

| **Column Name**    	| **Description**                                                                                   	|
|--------------------	|---------------------------------------------------------------------------------------------------	|
| longitude          	| Longitudinal location of block group of houses.                                                     |
| latitude           	| Latitudinal location of block group of houses                                                     	|
| housing_median_age 	| Median age of the houses in the block group                                                       	|
| total_rooms        	| Total rooms in the group of houses                                                                	|
| total_bedrooms     	| Total bedrooms in the block group                                                                 	|
| population         	| Total number of people in the block group                                                         	|
| households         	| Households comprised in the block group                                                           	|
| median_income      	| Median income calculated from the individual income of house.                                     	|
| median_house_value 	| Median house value of a block group.                                                              	|
| ocean_proximity    	| Indicating whether each block group is near the ocean, near the Bay Area, inland or on an island. 	|


## 4. Background on datasets

For our project we will be using the data set from kaggle [California Housing Prices](https://www.kaggle.com/datasets/camnugent/california-housing-prices)

The data contains information from the 1990 California census and pertains to houses found in a given California district. It contains one row per census block group. A block group is the smallest geographical unit for which the U.S. Census Bureau publishes sample data (a block group typically has a population of 600 to 3,000 people).


## 5. Research Question/Interest in Data Set

//Avi
//Kathryn

//TODO: A brief statement of the business, science, research, or personal interest you have in the data set which you hope to explore

```{r plot of data, echo=FALSE}
#TODO: if we want to go the extra mile, we can histogram or scatter plot our response variable without showing the code

```


## 6. Evidence of Loading into R

```{r read_data_file}
ca_housing_data = read.csv('../000_Data/california-housing-prices/housing.csv')
head(ca_housing_data)
```




```{r}
#TODO: load in the dataset and print first few obs
```


